Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.

To the extent that warming is coupled to drought, we expect that the tallest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.

What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2). 
Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].

Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), limited tree-ring evidence indicates that understory trees can exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].

Thus, in synthesis, warming temperatures will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, while much larger uncertainty remains surrounding the resoponses of understory trees.
Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].
When the canopy trees die, this will open canopy gaps, as discussed in the following section.

>>>>

As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
Our in-depth review sheds some light on how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.

To the extent that warming is coupled to drought, we expect that the tallest trees to usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that VPD in particular skews the stress more towards the exposed canopy trees.

It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
Leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to local climate but do not vary consistently with height (see section 4.2, Table 2). 
Beyond the optimal temperature, photosynthesis declines whereas respiration continues to increase, shifting the carbon balance from source to sink, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will reduce carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].

Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
<!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].

Thus, in synthesis, warming will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$, while greater uncertainty remains surrounding the resoponses of understory trees.
Both groups -- but probably disproportionately the upper-canopy trees -- are likely to experience increasing mortality, with patterns mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].
Death of canopy trees will open canopy gaps, as discussed in the following section.
